### 2012-04-13 trunk/ijcis-pos-model/sources/temporaldatabases.tex
new modifications
updated version.

When working with these time concepts, the Data Manipulation Language (\emph{DML}, which is part of the standard database querying language SQL) is extended to deal with possible temporal inconsistencies within the data and to handle more complex (temporal) queries. 
%Next to these concepts, also \textbf{user defined time} (\emph{UDT}) is discerned. UDT is an uninterpreted attribute in the date/time domain. This means that the attribute uses the date/time domain, but the database model does not treat the attribute differently from non-temporal attributes.
Depending on the time managed, a database is classified as either a \textbf{Valid Time Database} (\emph{VTDB}), a \textbf{Transaction Time Database} (\emph{TTDB}), a \textbf{bi-temporal database} (both valid and transaction time are managed) or a \textbf{tri-temporal database} (valid time, transaction time and decision time are managed).

@@

%%%%%%
% FUZZY VALIDITY PERIOD
%%%%%%
\vspace*{13pt}
\begin{center}
{
\includegraphics[scale=0.25]{./graphs/trapezoidalDistribution.pdf}

@@

\subsection{Understanding Valid-time Databases}
This subsection is devoted to describe the behaviour of a crisp valid-time database. For the shake of simplicity, only the three main operations in the Data Manipulation Language \emph{DML} a.k.a \emph{CRUD} (CReate Update, Delete) operations are shown. Usually the DML operations in a temporal database are re-defined e.g. a typical update sentence in SQL could be re-defined by means of several insert and update sentences. Therefore, in order to be clear, these high level primitives in the DMl for a valid-time database are usually noted as Insert, Modify and Delete. In the following subsections each primitive is defined and explained. Finally a illustrative example is given. For a more complete information on the behaviour of a bitemporal database, please refer to ~\cite{Jensen1994}.

@@

\subsubsection{Insert}
The interpretation of the insert operation is the following: The user wants to store in the database that an object $A$  is true for some period(s) of time $I$ in the given relation $R$. To indicate that the object $A$ is current in the relation, the special value Until Current, $\UC$  is used.

There are three main cases when performing a create operation:
\begin{enumerate}
\item The object $A$ was never in the relation $R$: The object is added with the valid-time indicated by the crisp interval $I$.

\item The object $A$ was part of some previously recorded state: A new version of the object is inserted in the relation $R$ with the valid-time indicated by $I$.

@@

\begin{align}
\label{eq:insert}
\INS \left(R ,A, I\right) =\\
\begin{cases}
\nonumber
R \cup \left(A, I \right), & \mbox{ if }  A \not \in R\\
R \cup  & \mbox{ if }  A \in R \wedge \\
R, & \mbox{ otherwise }  
\end{cases} 	
\end{align}

>>>>

When working with these time concepts, the Data Manipulation Language (\emph{DML}, which is part of the standard database querying language SQL) is extended to deal with possible temporal inconsistencies within the data and to handle more complex (temporal) queries. 
%Next to these concepts, also \textbf{user defined time} (\emph{UDT}) is discerned. UDT is an uninterpreted attribute in the date/time domain. This means that the attribute uses the date/time domain, but the database model does not treat the attribute differently from non-temporal attributes.
Depending on the time managed, a database is classified as either a \textbf{Valid Time Database} (\emph{VTDB}), a \textbf{Transaction Time Database} (\emph{TTDB}), a \textbf{bi-temporal database} (both valid and transaction time are managed) or a \textbf{tri-temporal} or \emph{multitemporal} database (valid time, transaction time and decision time are managed).

@@

%%%%%%
% FUZZY VALIDITY PERIOD
%%%%%%
\vspace*{13pt}
\begin{center}
{
\includegraphics[scale=0.25]{./graphs/comparisoncv.pdf}

@@

\subsection{Understanding Valid-time Databases}
This subsection is devoted to describe the behaviour of a crisp valid-time database. For the shake of simplicity, only the three main operations in the Data Manipulation Language \emph{DML} a.k.a \emph{CRUD} (CReate Update, Delete) operations are shown. Usually the DML operations in a temporal database are re-defined e.g. a typical update sentence in SQL could be re-defined by means of several insert and update sentences. Therefore, in order to be clear, these high level primitives in the DML for a valid-time database are usually noted as Insert, Modify and Delete. In the following subsections each primitive is defined and explained. Finally a illustrative example is given. For a more complete information on the behaviour of a bitemporal database, please refer to ~\cite{Jensen1994}.

@@

In order to simplify the algorithms for the manipulation of data, some auxiliary functions are defined:

\begin{definition}
\label{def:close-a-crisp-interval}
Consider two crisp intervals $I= \left[a ,b \right]$ and $J= \left[c ,d \right]$, $c > a$. The CloseL$\left(I, J\right)$ function allows to close the left-open interval $I$ with respect to the first value $d$ in $J$:
\begin{align}
\mbox{CloseL} \left( I, J \right) &=& \\ 
\begin{cases}
\nonumber
I & \mbox{ if } b \neq +\infty \\
I=\left[a, c \right[ & \mbox{ if } b = +\infty \wedge J > I
\end{cases}
\end{align}
\end{definition}

Analogously, it is defined the function to close an right open interval: 
\begin{definition}
\label{def:close-a-crisp-interval-r}
CloseR$\left(I, J\right)$ 
\begin{align}
\mbox{CloseR} \left( I, J \right) &=& \\ 
\begin{cases}
\nonumber
I & \mbox{ if } a \neq -\infty \\
I=\left]d, b \right[ & \mbox{ if } a = -\infty \wedge J < I
\end{cases}
\end{align}
\end{definition}

\begin{definition}
\label{def:current-in-relation}
Consider an object $A$ and a relation $R$ in a temporal database. The fucntion Current$\left(R, A \right)$ obtains the tuple $\left(A, I \right)$ that is current in the relation $R$, that is it, $I = \left[a, +\infty \right]$.

\begin{align}
\label{eq:current-in-relation}
\mbox{Current} \left(R, A \right) &=& \\ 
\begin{cases}
\nonumber
I & \mbox{ if } \exists b \in I : (A,I) \in R \wedge b = +\infty \\
\emptyset & \mbox{ in any other case }
\end{cases}
\end{align}
\end{definition}

Now it is possible to close the current version of an object by combining the two previously defined functions.

\begin{definition}
\label{def:close-current-version}
Consider an object $A$, the relation $R$ and a time interval $J$. Then, the function close-current$\left(R,A,J \right)$ closes any current version of the object $A$ if it exists and add the new version $\left(A, J \right)$.

\begin{eqnarray}
%\label{eq:close-current}
\text{Close-current} \left(R, A \right) =\\
\begin{cases}
\nonumber
\big \lbrace R - \left(A, I \right) \cup \left \lbrace \left(A, \mbox{ Close } \left(I, J\right) \right) \cup \left(A, J\right)\right \rbrace  \big \rbrace \\
\nonumber
\mbox{ if } I = \mbox{ Current } \left(R, A \right) \wedge J > \mbox{ Close } \left(I, J \right)   \\
\nonumber R , \text{ in any other case}
\end{cases}
\end{eqnarray}
\end{definition}

\subsubsection{Insert}
The user wants to store an object $A$ which is valid in the relation $R$ during the time interval specified by $I = \left(a, b \right)$.
%
%
%The interpretation of the insert operation is the following: The user wants to store in the database that an object $A$  is true for some period(s) of time $I$ in the given relation $R$. To indicate that the object $A$ is current in the relation, the special value Until Current, $\UC$  is used. 
There are three main cases when performing a create operation:
\begin{enumerate}
\item The object $A$ was never in the relation $R$: The object is added with the valid-time indicated by the crisp interval $I$.

\item The object $A$ is in the relation $R$, but it is not current

@@

\begin{align}
\label{eq:insert}
\INS \left(R ,A, I\right) =\\
\begin{cases}
\nonumber
R \cup \left(A, I \right), & \mbox{ if }  A \not \in R\\
R \cup  \left(A, I \right) & \mbox{ if }  A \in R \wedge \\
R, & \mbox{ otherwise }  
\end{cases} 	
\end{align}
